Using Incremental Many-to-One Queries to Build a Fast and Tight Heuristic for A* in Road Networks

### 2022-08-30 paper/ch_potentials.tex
wordsmithing

PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all vertices to one target vertex (or vice versa, the reverse case works analogously).
This is sometimes denoted as a \emph{all-to-one} (or \emph{one-to-all}) problem.
The preprocessing phase remains the same as for CH.
The query phase is split into two steps.
The first step is analogue to the CH query:
From $t$, all reachable vertices via reversed down-edges are explored as shown in Algorithm~\ref{algo:ch-backward}.
For the second step, PHAST utilizes an assignment of vertices into \emph{levels}.
These levels correspond to the importance ordering but allow multiple vertices in the same level.
However, no edge must connect two vertices within the same level.
Such a level assignment can be obtained by assigning all vertices without lower ranked neighbors to the lowest level.
All other vertices are then iteratively assigned to the next level above the highest level of any downward neighbor.

@@

To enable quick shortest path computations, we consider a two-phase setup with an additional offline preprocessing phase before the online query phase.
The input to the preprocessing phase is a lower bound graph $G_\ell = (V_{\ell}, E_{\ell})$ with lower bound weights $w_\ell$ where $w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
The output of the preprocessing is auxiliary data that allows to quickly compute distances on $G_\ell$ with respect to $w_\ell$.
In the applications considered in this paper, $w_\ell$ is always the \emph{freeflow} travel time.

@@

Avoiding tunnels and/or highways is a common feature of navigation devices.
Implementing this feature with CH-Potentials is easy.
We set $w_\ell$ to the freeflow travel time.
If an edge is a tunnel and/or a highway, we set $w_q$ to $+\infty$.
Otherwise, $w_q$ is set to the freeflow travel time.

@@

To support only live traffic, we set $w_\ell$ to the freeflow travel time.
$w_q$ is set to the travel time accounting for current traffic.
As traffic only increases the travel time along an edge, $w_\ell$ is a valid lower bound for $w_q$.
In a real world application, values from $w_q$ could be updated between queries.
This is all that is necessary to apply CH-Potentials in a live traffic scenario.

To combine live traffic with predicted traffic, we define a modified travel time function $w_q$ that is then used as query weights.
Denote by $w_p(e,\tau)$ the predicted travel time along edge $e$ at moment $\tau$.
Further, $w_c(e)$ is the travel time according to current live traffic.
Finally, we denote by $\tau_{\mathrm{soon}}$ the moment when we switch to predicted traffic.
In our experiments, we set $\tau_{\mathrm{soon}}$ to one hour in the future.
We need to make sure that the modified travel time function fulfills the FIFO property.
For this reason, we cannot make a hard switch at $\tau_{\mathrm{soon}}$.
Our modified travel time function linearly approaches the predicted travel time. % with a bounded slope.
%
Formally, we set $w_q(e,\tau)$ to $w_c(e)$, if $\tau \leq \tau_{\mathrm{soon}}$.
Otherwise, we check whether $w_p(e,\tau_{\mathrm{soon}}) < w_c(e)$ is true.
If it is the case, we set $w_q(e,\tau)$ to $\max\{w_c(e)+(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
Otherwise, we set $w_q(e,\tau)$ to $\min\{w_c(e)-(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
% We then run the algorithm of Section~\ref{sec:predicted-traffic} with $w_q$ as query weight.
In our implementation, we do not modify the representation of $w_p$ but evaluate the formulas above at each travel time evaluation.
We set $w_\ell$ to the freeflow travel time.

@@

\subsection{Inputs and Methodology}
Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map.\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}
To obtain the routing graph, we use the import from RoutingKit.\footnote{\url{https://github.com/RoutingKit/RoutingKit}}
The graph has 16M vertices and 35M edges.
For this instance, we have proprietary traffic data provided by Mapbox.\footnote{\url{https://mapbox.com}}
The data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the edge between the vertices.
It also includes traffic predictions for 38\% of the edges as predicted speeds for all five minute periods over the course of a week.
We exclude speed values which are faster than the freeflow speed computed by RoutingKit.
We also perform experiments on the Europe instance provided by PTV\footnote{\url{https://ptvgroup.com}} for the 9th DIMACS implementation challenge~\cite{DemetrescuGJ09}.
Additionally, we have three graphs with proprietary traffic predictions also provided by PTV.
The PTV instances are not OSM-based.
One is an old instance of Germany with traffic predictions from 2006 for 7\% of the edges which was used for evaluation of many competing time-dependent algorithms.
The other two are newer instances of Europe with predictions for 27\% (TDEur17) and 76\% (TDEur20) of the edges.
Table~\ref{tab:graphs} contains an overview over the instances.
In this table, we further report the sequential preprocessing running time to construct the augmented graphs and the space consumption.
We use the CH preprocessing from RoutingKit.\footnote{\url{https://github.com/RoutingKit/RoutingKit}}
For CCH, we use our own implementation with nested dissection orders obtained by InertialFlowCutter~\cite{GottesburenHUW19}.
We report preprocessing running times as averages over 10 runs.
To evaluate point-to-point queries, we generate 10\,000 queries where both source and target are vertices drawn uniformly at random and report average results.

@@

The performance of A* depends on the tightness of the heuristic and the overhead of evaluating the heuristic.
CH-Potentials computes optimal distance estimates with respect to $w_\ell$.
However, for most applications, there will be a gap between $w_q$ and $w_\ell$ (otherwise one could use CH without A*).
We evaluate the impact of the difference between $w_q$ and $w_\ell$ on the performance of A*.
The lower bound $w_\ell$ is set to the freeflow travel time.
The query weights $w_q$ are set to $\alpha \cdot w_\ell$, where $\alpha\ge 1$.
% When $\alpha = 1$, the heuristic is tight.
Increasing $\alpha$ degrades the heuristic's quality.
Figure~\ref{fig:scaled_weights} depicts the results.
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time grows quickly.
For $\alpha > 1.1$, the growth slows down.
This illustrates both the strengths and limits of our approach and goal-directed search in general.
CH-Potentials can only achieve competitive running times if the application allows for sufficiently tight lower bounds at preprocessing time.

>>>>

PHAST~\cite{dgnw-phast-13} is a CH extension that computes distances from all vertices to one target vertex (or vice versa, the reverse case works analogously).
This is sometimes denoted as a \emph{all-to-one} (or \emph{one-to-all}) problem.
The preprocessing phase remains the same as for CH.
The query phase is split into two steps.
The first step is analogue to the CH query:
From $t$, all reachable vertices via reversed down-edges are explored as shown in Algorithm~\ref{algo:ch-backward}.
For the second step, PHAST utilizes an assignment of vertices into \emph{levels}.
These levels correspond to the importance ordering but allow multiple vertices in the same level.
However, no edge must connect two vertices within the same level.
Such a level assignment can be obtained by assigning all vertices without lower-ranked neighbors to the lowest level.
All other vertices are then iteratively assigned to the next level above the highest level of any downward neighbor.

@@

To enable quick shortest path computations, we consider a two-phase setup with an additional offline preprocessing phase before the online query phase.
The input to the preprocessing phase is a lower bound graph $G_\ell = (V_{\ell}, E_{\ell})$ with lower bound weights $w_\ell$ where $w_\ell(e)$ must be a scalar value for every edge $e$ of $G_\ell$.
The output of the preprocessing is auxiliary data that allows to quickly compute distances on $G_\ell$ with respect to $w_\ell$.
In the applications considered in this paper, $w_\ell$ is always the \emph{free-flow} travel time.

@@

Avoiding tunnels and/or highways is a common feature of navigation devices.
Implementing this feature with CH-Potentials is easy.
We set $w_\ell$ to the free-flow travel time.
If an edge is a tunnel and/or a highway, we set $w_q$ to $+\infty$.
Otherwise, $w_q$ is set to the free-flow travel time.

@@

To support only live traffic, we set $w_\ell$ to the free-flow travel time.
$w_q$ is set to the travel time accounting for current traffic.
As traffic only increases the travel time along an edge, $w_\ell$ is a valid lower bound for $w_q$.
In a real world application, values from $w_q$ could be updated between queries.
This is all that is necessary to apply CH-Potentials in a live traffic scenario.

To combine live traffic with predicted traffic, we define a modified travel time function $w_q$ that is then used as query weights.
Denote by $w_p(e,\tau)$ the predicted travel time along edge $e$ at moment $\tau$.
Further, $w_c(e)$ is the travel time according to current live traffic.
Finally, we denote by $\tau_{\mathrm{soon}}$ the moment when we switch to predicted traffic.
In our experiments, we set $\tau_{\mathrm{soon}}$ to one hour in the future.
We need to make sure that the modified travel time function fulfills the FIFO property.
For this reason, we cannot make a hard switch at $\tau_{\mathrm{soon}}$.
Our modified travel time function linearly approaches the predicted travel time. % with a bounded slope.
%
Formally, we set $w_q(e,\tau)$ to $w_c(e)$, if $\tau \leq \tau_{\mathrm{soon}}$.
Otherwise, we check whether $w_p(e,\tau_{\mathrm{soon}}) < w_c(e)$.
If it is the case, we set $w_q(e,\tau)$ to $\max\{w_c(e)+(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
Otherwise, we set $w_q(e,\tau)$ to $\min\{w_c(e)-(\tau_{\mathrm{soon}}-\tau), w_p(e,\tau)\}$.
% We then run the algorithm of Section~\ref{sec:predicted-traffic} with $w_q$ as query weight.
In our implementation, we do not modify the representation of $w_p$ but evaluate the formulas above at each travel time evaluation.
We set $w_\ell$ to the free-flow travel time.

@@

\subsection{Inputs and Methodology}
Our main benchmark instance is a graph of the road network of Germany obtained from Open Street Map.\footnote{\url{https://download.geofabrik.de/europe/germany-200101.osm.pbf}}
To obtain the routing graph, we use the import from RoutingKit.\footnote{\url{https://github.com/RoutingKit/RoutingKit}}
The graph has 16M vertices and 35M edges.
For this instance, we have proprietary traffic data provided by Mapbox.\footnote{\url{https://mapbox.com}}
The data includes a live traffic snapshot from Friday 2019/08/02 afternoon and comes in the form of 320K OSM node pairs and live speeds for the edge between the vertices.
It also includes traffic predictions for 38\% of the edges as predicted speeds for all five minute periods over the course of a week.
We exclude speed values which are faster than the free-flow speed computed by RoutingKit.
We also perform experiments on the Europe instance provided by PTV\footnote{\url{https://ptvgroup.com}} for the 9th DIMACS implementation challenge~\cite{DemetrescuGJ09}.
Additionally, we have three graphs with proprietary traffic predictions also provided by PTV.
The PTV instances are not OSM-based.
One is an old instance of Germany with traffic predictions from 2006 for 7\% of the edges which was used for evaluation of many competing time-dependent algorithms.
The other two are newer instances of Europe with predictions for 27\% (TDEur17) and 76\% (TDEur20) of the edges.
Table~\ref{tab:graphs} contains an overview over the instances.
In this table, we further report the sequential preprocessing running time to construct the augmented graphs and the space consumption.
We use the CH preprocessing from RoutingKit.\footnote{\url{https://github.com/RoutingKit/RoutingKit}}
For CCH, we use our own implementation with nested dissection orders obtained by InertialFlowCutter~\cite{GottesburenHUW19}.
We report preprocessing running times as averages over 10 runs.
To evaluate point-to-point queries, we generate 10\,000 queries where both source and target are vertices drawn uniformly at random and report average results.

@@

The performance of A* depends on the tightness of the heuristic and the overhead of evaluating the heuristic.
CH-Potentials computes optimal distance estimates with respect to $w_\ell$.
However, for most applications, there will be a gap between $w_q$ and $w_\ell$ (otherwise one could use CH without A*).
We evaluate the impact of the difference between $w_q$ and $w_\ell$ on the performance of A*.
The lower bound $w_\ell$ is set to the free-flow travel time.
The query weights $w_q$ are set to $\alpha \cdot w_\ell$, where $\alpha\ge 1$.
% When $\alpha = 1$, the heuristic is tight.
Increasing $\alpha$ degrades the heuristic's quality.
Figure~\ref{fig:scaled_weights} depicts the results.
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time grows quickly.
For $\alpha > 1.1$, the growth slows down.
This illustrates both the strengths and limits of our approach and goal-directed search in general.
CH-Potentials can only achieve competitive running times if the application allows for sufficiently tight lower bounds at preprocessing time.
